---
title: "Predicting Retention"
author: "Jason Bryer, Ph.D."
date: "March 30, 2020"
output:
  ioslides_presentation: 
    smaller: yes
    widescreen: yes
editor_options: 
  chunk_output_type: console
---

<style>
blockquote {
  background: #f9f9f9;
  border-left: 5px solid #ccc;
  margin: 1.5em 10px;
  padding: 0.5em 1.5em;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
library(DT)
library(ggplot2)
library(psych)
```

## Overview

In higher education, being able to predict retention is important for multiple reasons including:

* Provide support services for students "at-risk"
* Used by some college rankings (e.g. [US News](https://www.usnews.com/best-colleges))

Many colleges use admissions tests (e.g. ACT, SAT) to determine admission to the college. Our guiding research question is:

**Does addmission tests predict college retention?**

Retention is defined as:

> A measure of the rate at which students persist in their educational program at an institution, expressed as a percentage. For four-year institutions, this is the percentage of first-time bachelors (or equivalent) degree-seeking undergraduates from the previous fall who are again enrolled in the current fall. For all other institutions this is the percentage of first-time degree/certificate-seeking students from the previous fall who either re-enrolled or successfully completed their program by the current fall. ([IPEDS, 2019](https://surveys.nces.ed.gov/ipeds/VisInstructions.aspx?survey=6&id=30074&show=all))

## Data Source

The [Integrated Postsecondary Education Data System (IPEDS)](https://nces.ed.gov/ipeds/) provides information about all higher education institutions that provide Federal Finacial Aid to students. 

```{r, message=FALSE, echo=FALSE}
library(ipeds)
data(surveys)
DT::datatable(surveys[,1:3], rownames = FALSE)
```

## Data Preparation

The [`ipeds`](https://github.com/jbryer/ipeds) R package provides an interface to download IPEDS data directly into R.

```{r message=FALSE, warning=FALSE, cache=TRUE}
directory <- getIPEDSSurvey('HD', 2011)
admissions <- getIPEDSSurvey("IC", 2011)
retention <- getIPEDSSurvey("EFD", 2011)
```

We will subset the columns we are interested in and rename them.

```{r}
directory <- directory[,c('unitid', 'instnm', 'sector', 'control')]
admissions <- admissions[,c('unitid', 'admcon1', 'admcon2', 'admcon7', 'applcnm', 
						   'applcnw', 'applcn', 'admssnm', 'admssnw', 'admssn', 
						   'enrlftm', 'enrlftw', 'enrlptm', 'enrlptw', 'enrlt', 
						   'satnum', 'satpct', 'actnum', 'actpct', 'satvr25', 
						   'satvr75', 'satmt25', 'satmt75', 'satwr25', 'satwr75', 
						   'actcm25', 'actcm75', 'acten25', 'acten75', 'actmt25', 
						   'actmt75', 'actwr25', 'actwr75')]
retention <- retention[,c('unitid', 'ret_pcf', 'ret_pcp')]
```

## Data Preparation: Rename Columns

```{r}
names(admissions) <- c("unitid", "UseHSGPA", "UseHSRank", "UseAdmissionTestScores", 
					  "ApplicantsMen", "ApplicantsWomen", "ApplicantsTotal", 
					  "AdmissionsMen", "AdmissionsWomen", "AdmissionsTotal", 
					  "EnrolledFullTimeMen", "EnrolledFullTimeWomen", 
					  "EnrolledPartTimeMen", "EnrolledPartTimeWomen", 
					  "EnrolledTotal", "NumSATScores", "PercentSATScores", 
					  "NumACTScores", "PercentACTScores", "SATReading25", 
					  "SATReading75", "SATMath25", "SATMath75", "SATWriting25", 
					  "SATWriting75", "ACTComposite25", "ACTComposite75", 
					  "ACTEnglish25", "ACTEnglish75", "ACTMath25", "ACTMath75", 
					  "ACTWriting25", "ACTWriting75")
names(retention) = c("unitid", "FullTimeRetentionRate", "PartTimeRetentionRate")

```

## Data Preparation: Recoding

Recode the `openadmission` and `distanceEd` variables to factors and `enrollment` to an integer.

```{r message=FALSE, warning=FALSE}
admissionsLabels = c("Required", "Recommended", "Neither requiered nor recommended", 
					 "Do not know", "Not reported", "Not applicable")
admissions$UseHSGPA = factor(admissions$UseHSGPA, levels=c(1,2,3,4,-1,-2), 
							labels=admissionsLabels)
admissions$UseHSRank = factor(admissions$UseHSRank, levels=c(1,2,3,4,-1,-2), 
							labels=admissionsLabels)
admissions$UseAdmissionTestScores = factor(admissions$UseAdmissionTestScores, levels=c(1,2,3,4,-1,-2), 
							labels=admissionsLabels)
```


## Data Preparation: Merging

```{r}
ret <- merge(directory, admissions, by="unitid")
ret <- merge(ret, retention, by="unitid")
#Use schools that require or recommend admission tests
ret2 <- ret[ret$UseAdmissionTestScores %in% 
		   	c('Required', 'Recommended', 'Neither requiered nor recommended'),] 
#Remove schools with low retention rates. Are these errors in the data?
ret2 <- ret2[-which(ret2$FullTimeRetentionRate < 20),] 
```

```{r}
head(ret2, n = 3)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE}
DT::datatable(ret2)
```

## Data Preparation: SAT and ACT Scores

IPEDS only provides the 25th and 75th percentile in SAT and ACT scores. We will use the mean of these two values as a proxy for the mean.

```{r, warning=FALSE, message=FALSE}
ret2$SATMath75 <- as.numeric(ret2$SATMath75)
ret2$SATMath25 <- as.numeric(ret2$SATMath25)
ret2$SATMath <- (ret2$SATMath75 + ret2$SATMath25) / 2
ret2$SATWriting75 <- as.numeric(ret2$SATWriting75)
ret2$SATWriting25 <- as.numeric(ret2$SATWriting25)
ret2$SATWriting <- (ret2$SATWriting75 + ret2$SATWriting25) / 2
ret2$SATTotal <- ret2$SATMath + ret2$SATWriting
ret2$NumSATScores <- as.integer(ret2$NumSATScores)
```

## Data Preparation: Selectivity

Calculate the the selectivity of the institution by calculating the acceptance rate (i.e. # admissions / # applicants).

```{r, warning=FALSE, message=FALSE}
ret2$Selectivity <- as.numeric(ret2$AdmissionsTotal) / 
				as.numeric(ret2$ApplicantsTotal)
ret2$UseAdmissionTestScores <- as.factor(as.character(ret2$UseAdmissionTestScores))
```

## Descriptive Statistics: FullTimeRetentionRate

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3.5, fig.align='center'}
psych::describe(ret2$FullTimeRetentionRate)
ggplot(ret2, aes(x = FullTimeRetentionRate)) + geom_histogram()
```

## Descriptive Statistics: SATMath

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3.5, fig.align='center'}
psych::describe(ret2$SATMath)
ggplot(ret2, aes(x = SATMath)) + geom_histogram()
```

## Descriptive Statistics: SATWriting

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3.5, fig.align='center'}
psych::describe(ret2$SATWriting)
ggplot(ret2, aes(x = SATWriting)) + geom_histogram()
```

## SAT Scores vs. Full-Time Retention

```{r echo=FALSE, fig.height=5, fig.width=10, message=FALSE, warning=FALSE}
ggplot(ret2, aes(x=SATTotal, y=FullTimeRetentionRate, size=NumSATScores, 
				 color=UseAdmissionTestScores)) + 
	geom_point(alpha = 0.3) + 
	geom_smooth(method='loess', alpha=0.1) +
	xlab('Median SAT Score') + ylab('Full-Time Retention Rate') +
	scale_size('Number of Scores Reported') +
	scale_color_brewer('Use Admission Test Scores', palette = 2, type = 'qual')
```

# Regression Analysis

## Regression Results

```{r}
lm.out <- lm(FullTimeRetentionRate ~ SATWriting + SATMath + 
			 Selectivity + UseAdmissionTestScores, 
			 data=ret2, 
			 weights=ret2$NumSATScores)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
options('width' = 120)
lm.out.sum <- summary(lm.out)
printCoefmat(lm.out.sum$coefficients)
```

Adjusted $R^2$ = `r round(lm.out.sum$adj.r.squared, digits = 2)`


## Conclusion (Regression)

SAT scores are a significant predictor of full-time retention rates for higher education institutions in the United States.

SAT scores and selectivity account for approximately `r round(round(lm.out.sum$adj.r.squared * 100))`% of the variance of full-time retention.

# Null Hypothesis Test

## Descriptive Statistics: UseAdmissionTestScores

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3.5, fig.align='center'}
table(ret2$UseAdmissionTestScores)
ggplot(ret2, aes(x = UseAdmissionTestScores)) + geom_bar()
```


## Null Hypothesis Test

$H_0$: There is no difference in retention rate between institutions that require SAT scores and those that do not.

$H_A$: There is a difference in the retention rate between institutions that require SAT scores and those that do not.

We will exclude the "Recommended" group so that we have two groups.

```{r}
ret3 <- ret2[ret2$UseAdmissionTestScores != 'Recommended',]
```

## Student's t-test

```{r}
t.results <- t.test(FullTimeRetentionRate ~ UseAdmissionTestScores, data = ret3)
print(t.results)
```

## Conclusion (Null Hypothesis test)

There is sufficient evidence to reject the null hypothesis (*p* < 0.01). There is a difference in retention rates between schools that require SAT scores (`r round(t.results$estimate[2], digits=2)`) and those that do not (`r round(t.results$estimate[1], digits=2)`).

We are 95% confident the true difference is between `r round(t.results$conf.int[1], digits = 2)` and `r round(t.results$conf.int[2], digits = 2)`.

# ANOVA

## Null Hypothesis for ANOVA

$H_0$: The mean retention rate is the same whether an institution requires, recommends, and neither require nor recommend SAT scores.

$H_A$: There is a difference in the mean retention scores between the three groups.

## ANOVA

```{r}
anova.results <- aov(FullTimeRetentionRate ~ UseAdmissionTestScores, data = ret2)
summary(anova.results)
```

### Conclusion (ANOVA)

There is sufficient evidence to reject the null hypothesis (p < 0.01), there is a difference in the retention rate depending on the requirement of SAT scores.



